---
title: "Homework 2"
author: August Jonasson
date: "`r Sys.Date()`"
output: github_document
---
```{r, include=FALSE}
library(tidyverse, lubridate)
library(scales)
library(knitr)
```

## Exercise 1: Apartment prices

We start by loading the data into a tibble.
```{r, message=FALSE}
# creating a dataframe containing the apartment prices
apartments <- read_csv("../hw_data/Booli_sold.csv")
```

#### Task a

Illustrating how the selling price is dependent on the living area using a
scatter-plot.
```{r, warning=FALSE}
# scatter-plot of selling price (in millions) and living area
apartments %>%
  ggplot(aes(x=livingArea, y=soldPrice/1000000)) +
  geom_point() +
  xlab("Living area (m^2)") +
  ylab("Selling price (in millions)") +
  ggtitle("Apartment selling price and living area in Ekhagen 2012/10-2018/11")
```

#### Task b

Now, lets examine the price per square meter. We use a scatter-plot 
combined with a regression line to illustrate the trend over time.
```{r, message=FALSE, warning=FALSE}
# creating new variable ppm2 (price per m^2 (in thousands)) and then creating a 
# scatter plot overlaid with a regression line, to illustrate the trend
apartments %>%
  mutate(ppm2 = soldPrice / livingArea / 1000) %>%
  ggplot(aes(x=soldDate, y=ppm2)) +
  geom_point() +
  geom_smooth(method = "loess") +
  scale_x_date(date_breaks = "1 year", labels = date_format(format = "%Y")) +
  xlab("Time of sale (date)") +
  ylab("Price per m^2 (in thousands)") +
  ggtitle("Price per m^2 trend in Ekhagen")
  
```

#### Task c

Let's make a table of number of rooms in relation to construction year, of the 
apartments from our data. The column headers are the number of rooms, and the
rows represent the construction year.

```{r, message=FALSE}
# using kable to make the table slightly prettier
table <- table(apartments$constructionYear, apartments$rooms)
kable(table, caption = "Table 1", "simple")
```

#### Task d

It would be interesting to see if the living area of the sold apartments
indicate some type of distribution.
```{r, warning=FALSE, message=FALSE}
# plotting a histogram of living area, of the sold apartments
apartments %>%
  ggplot(aes(x=livingArea)) +
  geom_histogram()
```

Pretty hard to tell. Maybe because family sizes differ. Unfortunately, we don't
have that data.

#### Task e

Lets plot the price per square meter (in thousands) for 1, 2, 3, and 4-room 
apartments using box plot.
```{r message=FALSE, warning=FALSE}
# creating box plots for differently roomed apartments and the their m^2-price
apartments %>%
  filter(rooms %in% c(1,2,3,4)) %>%
  mutate(ppm2 = soldPrice / livingArea / 1000) %>%
  ggplot(aes(x = as.factor(rooms), y=ppm2)) +
  geom_boxplot() +
  xlab("Apartment-rooms (kitchen and bathroom excluded)") +
  ylab("Price per m^2 (in thousands)") +
  ggtitle("Price per m^2 for differently roomed apartments in Ekhagen")
```


## Exercise 2: Folkhälsomyndigheten COVID cases, and excel

### Data wrangling tasks

#### Task a

The file consists of daily confirmed cases, resulting casualties, and patients
in need of intensive care, due to COVID, in Sweden. This information is enriched
by factors such as province, municipality and patient age and sex. All of the
sheets are different manipulations of this data.
  
    Sheet 1: "Antal per dag region" - confirmed cases by province, per day
    Sheet 2: "Antal avlidna per dag" - deceased patients per day
    Sheet 3: "Antal intensivvårdade per dag" - ICU-patients per day
    Sheet 4: "Totalt antal per region" - total count of Sheet 1, 2 and 3, by province
    Sheet 5: "Totalt antal per kön" - total count of Sheet 1, 2 and 3, by sex
    Sheet 6: "Totalt antal per åldersgrupp" - total count of Sheet 1, 2 and 3, by age
    Sheet 7: "Veckodata region" - weekly data with a cumulative aspect, by province
    Sheet 8: "Veckodata Kommun_stadsdel" - weekly data with a cumulative aspect, by municipality
    Sheet 9: "FOHM 30 Sep 2020" - Information about how often the data is updated

#### Task b

```{r, include=FALSE}
library(readxl)
```


Lets read all of the sheets in as separate tibbles. Sheet 9 is ignored since it 
doesn't contain anything useful.
```{r, warning=FALSE}
# excel file path
file <- "../hw_data/Folkhalsomyndigheten_Covid19.xlsx"

# Sheet 1 = s1, Sheet 2 = s2...
s1 <- read_excel(file, sheet = 1) # Antal per dag region
s2 <- read_excel(file, sheet = 2) # Antal avlidna per dag
s3 <- read_excel(file, sheet = 3) # Antal intensivvårdade per dag
s4 <- read_excel(file, sheet = 4) # Totalt antal per region
s5 <- read_excel(file, sheet = 5) # Totalt antal per kön
s6 <- read_excel(file, sheet = 6) # Totalt antal per åldersgrupp
s7 <- read_excel(file, sheet = 7) # Veckodata region
s8 <- read_excel(file, sheet = 8) # Veckodata Kommun_stadsdel
```

#### Task c

Printing the requested table containing the first and last five rows of the s2-
tibble.

```{r}
# using kable to print the table
s2_head <- head(s2, 5)
s2_tail <- tail(s2, 5)
kable(rbind(s2_head, s2_tail), caption = "Table 2", "simple")

```

From Table 2, we can see that the last entry seems a bit suspicious. "Uppgift
saknas" is Swedish for "NA". Upon further investigation we can tell that the
variable "Datum_avliden" is actually parsed as chr - a string, but this is
supposed to be a date. We redefine the read function for this tibble to not
include the last row, and then check the class of the variable again.

```{r}
s2 <- read_excel(file, sheet = 2, n_max = 203)
class(s2$Datum_avliden)
```

Now it's parsed as a date and we are good.

#### Task d

```{r}
class(s8$Stadsdel)
```
As we can tell from this result, the 'Stadsdel' variable from Sheet 8 has the type
'logical'. Using the 'unique' function we can also tell that 'NA' is the only
value that appears under this variable, but a quick check in the original excel
file reveals that this shouldn't be the case. The first 1882 rows should have
the value 'NA', but then some strings should appear. 

Reading up on the read_excel documentation we find out that the reason this
happens is because the guess_max argument, by default, is set to 1000, ergo the
read_excel function makes up its mind about the data type based on the first 
1000 rows. Since our first 1000 rows were 'NA' (read as FALSE?), it decided
that the data type should be 'logical'. This can easily be fixed by increasing 
the guess_max value.

```{r}
s8 <- read_excel(file, sheet = 8, guess_max = 1883)
class(s8$Stadsdel)
```
Now it's parsed as a string and we are good.

#### Task e

Checking the values for the 'tot_antal_fall' and 'nya_fall_vecka' variables in 
Sheet 8.
```{r}
class(s8$tot_antal_fall)
class(s8$nya_fall_vecka)
```
They are both parsed as strings, but from the excel file it looks like they
should be numerical.

From looking further at the excel file we see that there are several values
in both columns typed in as '<15', instead of the precise value of that day.
Some of the '<15' values we can actually deduce to being a precise number,
based on the days that follow, but for most of them, they are as good as 'NA' -
missing values. Since we are not paid to do this we take the very crude approach
and replace all of them with 'NA'. This can easily be done using the as.numeric
function, since this automatically assigns 'NA' to all values that can't be
parsed as numbers.

```{r, warning=FALSE}
# creating a modified version of s8, replacing all '<15' with 'NA'
s8_mod <- s8 %>%
  mutate(tot_antal_fall = as.numeric(tot_antal_fall),
         nya_fall_vecka = as.numeric(nya_fall_vecka))

class(s8_mod$tot_antal_fall)
class(s8_mod$nya_fall_vecka)
```
Now they both have the numeric type, and we are good.

### Statistics and plotting

#### Task a

We want to create a new data frame containing the total number of confirmed
cases in each region, and country-wise.
```{r}
# the new data frame is created using summarize and across, with the sum fctn.
total_by_province <- s1 %>%
  summarize(across(2:last_col(), ~ sum(.)))
```

Now, to extract the total number of cases and the provinces with the most and
least cases respectively.
```{r}
# we first transpose the data frame since it's kinda annoying to use horizontally
totprov_trans <- data.frame(t(total_by_province)) %>%
  rename(antal_fall = t.total_by_province.) %>%
  arrange(desc(antal_fall))

# display total, max and min using kable
kable(rbind(head(totprov_trans, 2), tail(totprov_trans, 1)),
      caption = "Table 3: Total cases, max and min by provinces",
      "simple")
```

As we can see from Table 3, the total number of cases was 92863, the province
with the most cases was Stockholm, and the province with the least cases was 
Gotland.

While the provincial counts on their own are interesting, comparing them to
each other might be a bit misleading. Partly because the population sizes differ
greatly. Additionally, the population densities probably has some effect as well
, since we are talking about the spread of a decease that's transmitted through
peoples breath. If one were to compare provinces, these factors (and probably
many more) should be considered.

#### Task b

Plotting the daily number of reported cases nationwide in Sweden.
```{r}
s1 %>%
  filter(Statistikdatum > as.POSIXct("2020-03-15")) %>%
  ggplot(aes(x=Statistikdatum, y=Totalt_antal_fall)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 month", labels = date_format("%m-%d")) +
  xlab("Date (month-day)") +
  ylab("Reported cases") +
  ggtitle("Sweden daily reported cases of COVID during 2020")
```

And for the cumulative value of the reported cases.
```{r}
s1 %>%
  filter(Statistikdatum > as.POSIXct("2020-03-15")) %>%
  mutate(kumulativt_antal = cumsum(Totalt_antal_fall)) %>%
  ggplot(aes(x=Statistikdatum, y=kumulativt_antal)) +
  geom_line() +
  scale_x_datetime(date_breaks = "1 month", labels = date_format("%m-%d")) +
  xlab("Date (month-day)") +
  ylab("Reported cases") +
  ggtitle("Sweden cumulative reported cases of COVID during 2020")
```

#### Task c

Deriving the total number of weekly reported cases in Sweden and plotting as a
bar plot.
```{r}
s7 %>%
  group_by(veckonummer) %>%
  summarize(tot_antal_fall_vecka = sum(Antal_fall_vecka)) %>%
  ggplot(aes(x=as.factor(veckonummer), y=tot_antal_fall_vecka)) +
  geom_col() +
  xlab("Week number") +
  ylab("Reported cases") +
  ggtitle("Weekly reported cases of COVID in Sweden 2020")
```

