---
title: The Right Spotify Playlist for the Occasion
author: August Jonasson (@aggelitoo)
date: 2023-01-10
output:
  html_document:
    self_contained: true
    toc: true
    toc_depth: 2
---

# Abstract

In this report we go use RSelenium to scrape the most popular tracks in the
world for the year 2022, from the Spotify top charts.
We then enrich the data with audio features such as valence, energy
and tempo using the Spotofy API and the R wrapper spotifyr. In the end, we use
these  variables in combination with the other data in order to automatically
generate two playlists, according to some emotion we want to facilitate.
We believe one of the playlists was a success, and the other one not so much,
and we believe the reason for this is that the second playlist wasn't
geographically restricted enough, while the first one was.

# Introduction

A lot of us use music in order to facilitate an emotion or make a task easier
to perform. When we have gone through a difficult break-up we look for music
that validates our sad feelings, and when we decide to have a pizza baking Friday
night, we want the music that sets the perfect atmosphere for kneading that
dough like a true Italian master.

Often times, however, we have to spend a lot of time
finding the right music, and creating the playlists that we require - and then
also having to re-create playlists that we have grown tired of. Wouldn't it be
nice if there was a way to automatically create the playlist you are looking for,
while at the same time ensuring some aspect of novelty?

Well, that might not be that far off (it is already somewhat here)!
In this report we will make a brief
exploration into this subject through the use of Spotify. We will first seek
out the most popular music (on Spotify) from around the globe, for the year 2022,
and then enrich
the tracks with features provided by the Spotify API. These feataures
include several different parameters, such as danceability, valence, energy and
tempo, just to name the ones focused solely on the "Mood" of the track. In this
project we will focus solely on the valence, energy and tempo parameters in order
to create playlists. Through different combinations of valence and energy, one
should be able to mimic emotions such as happy, sad, angry and relaxed. The
tempo variable will be used to make the playlists feel coherent. More on these
variables later in the report.

The full spectrum of the spotify audio features can be found [here](https://developer.spotify.com/discover/#audio-features-analysis).

This report is intended as a guide/introduction into the world of generating
playlists automatically through the use of R, web scraping spotify weekly charts
using RSelenium, and the Spotify for Developers API using the R wrapper spotifyr.
A lot of emphasis will be put on the actual gathering of the data, since regular
web scraping doesn't work on the Spotify charts. When the data has been gathered,
some very basic examples of playlist building will be shown.

As such, the reader should be familiar with R Tidyverse and, perhaps,
like listening to music. An interest in psychology and affect models will
definitely elevate the experience as well.

# Data acquisition

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(lubridate)
library(spotifyr)
library(rvest)
library(RSelenium)
```

First step is to scrape historical data from the Spotify charts URLs. These
have to be gathered per week, per country, in order to build a larger data frame
consisting of all countries and their top songs over the whole year. This will
be one of the more challenging parts of this project, since the site is guarded
behind a login page, and the use of rvest in order to get through this page
isn't possible. We therefor have to use RSelenium in order to create a script
that automates the gathering of the data through a regular web browser. The
current documentation for RSelenium can be found
[here](https://www.rdocumentation.org/packages/RSelenium/versions/1.7.9).

There are csv files available on the Spotify charts website, but only per week
and country. Therefore, we want to automate the downloading of these csv files,
and then create one large data frame out of the ~3500 csv files that we gather.

First, lets create the vectors that we need in order to iterate through
the spotify charts URLs.
```{r}
# these variables will be used for both scraping and csv_reading

# all the week-ends of 2022 (Jan - Nov)
date_range <- seq(ymd('2022-01-06'),ymd('2022-12-01'), by = '1 week')

# all countries available on spotify charts, by two letter country code
countries = c("ar", "au", "at", "by", "be", "bo", "br", "bg", "ca", "cl", "co",
              "cr", "cy", "cz", "dk", "do", "ec", "eg", "sv", "ee", "fi", "fr",
              "de", "gr", "gt", "hn", "hk", "hu", "is", "in", "id", "ie", "il",
              "it", "jp", "kz", "lv", "lt", "lu", "my", "mx", "ma", "nl", "nz",
              "ni", "ng", "no", "pk", "pa", "py", "pe", "ph", "pl", "pt", "ro",
              "sa", "sg", "sk", "za", "kr", "es", "se", "ch", "tw", "th", "tr",
              "ae", "ua", "gb", "uy", "us", "ve", "vn")
```

Now, we can run the following script in order to gather all of the csv files.
One initial, manual, login to the Spotify charts website has to performed with
the Spotify account that should be used, in order to accept the terms of the
site. [This](https://charts.spotify.com) is the Spotify charts site. The
following code chunk automates over 10 thousand clicks that are needed to download
the data and takes several hours to run.
```{r, message=FALSE, eval=FALSE}
# Spotify account credentials
user <- "studentgarbage@gmail.com"  # this account was created for this project
password <- "Losenord1234567890"    # and doesn't contain any sensitive information

# setting up the browser to be used in the script
rD <- rsDriver(browser = "firefox", chromever = NULL)
remDr <- rD[['client']]
remDr$navigate("https://accounts.spotify.com/sv/login?continue=https%3A%2F%2Fcharts.spotify.com/login")

# logging in to the spotify charts website, in order to access the charts
remDr$findElement(using = "id", value = "login-username")$sendKeysToElement(list(user))
remDr$findElement(using = "id", value = "login-password")$sendKeysToElement(list(password))
remDr$findElement(using = "id", value = "login-button")$clickElement()

# clicking accept cookies popup
Sys.sleep(3)
remDr$findElement(using = "id", value = "onetrust-accept-btn-handler")$clickElement()

# function to download all the weekly charts for each week, for each country
# we use 'possibly' to wrap function since not all weekly charts are available
# files are downloaded to whatever default downloaded files folder
map(countries, function(x){
      map(date_range, possibly(function(y) {
        # navigating to the week/country site
        remDr$navigate(str_c("https://charts.spotify.com/charts/view/regional-",
                             x, "-weekly/", as.character(y)))
        Sys.sleep(2)
        # finding the download csv file button and clicking it
        remDr$findElement(using = "xpath",
                          value = "/html/body/div/div/div/main/div[2]/div[3]/div/div/a/button")$clickElement()
      }, otherwise = NULL))
})
```

Now, that we have gathered all of the weekly charts for 2022, from around the
world, as csv files - we can build the actual data frame. We then save it as an
RData file in order to avoid having to re-create it on each run.
```{r, eval=FALSE, message=FALSE}
# ca 3500 csv files, 200 rows per file - takes some time to compile
charts_df <- map_df(countries, function(x){
  map(date_range, possibly(function(y) {
    read_csv(str_c("spotify_csv_files/regional-", x, "-weekly-", as.character(y), ".csv")) %>%
      mutate(country = x, week_end = y)
  }, otherwise = NULL))
})

# then saving the data frame as .RData to avoid having to rerun this code
save(charts_df, file = "project_RData/charts.RData")
```

Second step is to use the Spotify API in order to retrieve the "Mood"
variables (valence, energy and tempo), of the tracks, and join
this with the data frame we just created. We use the R wrapper spotifyr to do
this.

In order to perform this step, one has to create a Spotify for Developers
account (regular spotify account can be used). Otherwise, an access token can't
be fetched, and the Spotify API won't work
([Spotify for Developers](https://developer.spotify.com/dashboard/login)).
Create an app on the developer dashboard and find its credentials (id and secret).

Since the API only allows for 100 calls at a time, when fetching track features,
we have to first identify all of the unique songs and then break them up into
chunks of 100. We can then proceed with the fetching of the features.
```{r, echo=FALSE, message=FALSE}
# authentication, in order to start fetching feature data from spotify api
# id and secret are spotify for developers project credentials
id <- 'ac4d8e6092014430ad715955013b36ea'
secret <- '9c39a489ca794a6d8bfc68e98b742d60'
Sys.setenv(SPOTIFY_CLIENT_ID = id)
Sys.setenv(SPOTIFY_CLIENT_SECRET = secret)
access_token <- get_spotify_access_token()
```


```{r, cache=TRUE, eval=FALSE}
# loading the data
load("project_RData/charts.RData")

# create a vector of songs that appear in the df, by their unique URI
unique_tracks <- charts_df %>%
  pull(uri) %>%
  unique() %>%
  str_remove(pattern = "spotify:track:") # to get Track ID, to use spotifyr

# breaking up tracks into chunks of 100
chunk_length <- 100
chunked_unique_tracks <- c(split(unique_tracks, ceiling(seq_along(unique_tracks) / chunk_length)))

# calling 100 songs at a time from the api to get features for all songs
features_df <- map_dfr(chunked_unique_tracks, ~ get_track_audio_features(.x))

# we also might want to fortify the data by adding country names and continents
# check sources for url to this data
country_and_continent_codes <- read_csv("project_RData/country_and_continent_codes.csv") %>%
  mutate(Two_Letter_Country_Code = tolower(Two_Letter_Country_Code))

# we can now finally create a complete data frame, with tracks, features, week
# dates when the songs were popular, and country and continent
charts_features_df <- charts_df %>%
  left_join(country_and_continent_codes, by = c("country" = "Two_Letter_Country_Code")) %>%
  left_join(features_df, by = "uri")

# saving this data frame as well, in order to avoid having to call API each time
save(charts_features_df, file = "project_RData/charts_features.RData")
```

```{r, echo=FALSE}
load("project_RData/charts_features.RData")
head(charts_features_df, 5)
```
We now finally have everything that we need in order to start experimenting
with the mood variables and create some playlists.

# Creating some playlists

This is a good spot to explain the Mood variables a bit more in dept. These are
the descriptions that Spotify give in their API documumentation.

 - Valence: A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).\
 - Energy: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.\
 - Tempo: The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration.

Now, if one were to examine the energy/valence relationship visually, with
valence on the x-axis and energy on the y-axis (centered around zero),
a psychologist might be eager to
say that observations further into the first quadrant represent feelings of
happiness and delight, the second quadrant would be anger and frustration, the
third sadness and depression, and the fourth would finally be something akin to
feeling relaxed and content (to name them very roughly).

Here are some example tracks, shown in this context, that you might recognize,
or be able to imagine what they sound like.

```{r, message=FALSE, echo=FALSE}
example_songs <- charts_features_df %>%
  select(track_name, artist_names, valence, energy, tempo, uri) %>%
  mutate(valence = valence - 0.5,
         energy = energy - 0.5) %>%
  mutate(track_artist = str_c(track_name, artist_names, sep = " - ")) %>%
  distinct() %>%
  filter((track_name == "Dancing Queen") | (track_name == "Hive Mind") |
           (track_name == "Something In The Way") | track_name == "Africa")

example_plot <- example_songs %>%
  ggplot(aes(x=valence, y=energy, shape = track_artist)) +
  geom_point(size = 4) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  xlim(-0.5, 0.5) +
  ylim(-0.5, 0.5) +
  geom_label(x=-0.3, y=0.32, label = "Angry/Turbulent", color = "red") +
  geom_label(x=-0.3, y=-0.32, label = "Sad/Depressed", color = "blue") +
  geom_label(x=0.3, y=0.32, label = "Happy/Delighted", color = "green") +
  geom_label(x=0.3, y=-0.32, label = "Relaxed/Mellow", color = "purple") +
  ggtitle("Valence/energy relationship of some popular songs")

example_plot
```

Let's set a goal for this project to create two playlists. One slow sad/depressed
playlist from the Nordic countries' charts, and one happy and upbeat playlist
pulling tracks from the entire world.

When we create the sad playlist, we only use songs with a tempo of below 100 bpm,
since songs faster than this might go against the feeling we are looking for.

```{r}
nordic_sad <- charts_features_df %>%
  # only the nordic charts for 2022
  filter(country %in% c("se", "no", "fi", "dk", "is")) %>%
  select(track_name, artist_names, valence, energy, tempo, uri) %>%
  mutate(valence = valence - 0.5, energy = energy - 0.5) %>%
  distinct() %>% # don't want duplicate songs
  # getting some, allegedly, sad songs - based on tempo as well
  filter(valence < -0.2 & energy < -0.1 & tempo < 100)

head(nordic_sad, 5)
```

These are the resulting tracks visualized on the valence/energy plot.

```{r, echo=FALSE}
nordic_sad_plot <- nordic_sad %>%
  ggplot(aes(x=valence, y=energy)) +
  geom_point(size = 0.3) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  xlim(-0.5, 0.5) +
  ylim(-0.5, 0.5) +
  geom_label(x=-0.3, y=0.32, label = "Angry/Turbulent", color = "red") +
  geom_label(x=-0.3, y=-0.32, label = "Sad/Depressed", color = "blue") +
  geom_label(x=0.3, y=0.32, label = "Happy/Delighted", color = "green") +
  geom_label(x=0.3, y=-0.32, label = "Relaxed/Mellow", color = "purple") +
  ggtitle("(Allegedly) Sad songs popular in the Nordic countries 2022")

nordic_sad_plot
```

Now, we can pick out a playlist. We are doing this using a set seed in order to
keep the sampling from these songs consistent. If we are not happy with the
playlist, or, if we grow tired of it - we can just change the seed. A good
playlist length might around 25 songs, so let's use this.

Lets create a playlist using spotifyr.
More details on the spotifyr package can be found
[here](https://www.rdocumentation.org/packages/spotifyr/versions/2.1.1).

```{r, message=FALSE, eval=FALSE}
# create the actual playlist using spotifyr
create_playlist(user_id = "31m5lue5uif6evna5thejrywzzgi",
                name = "nordic_sad")
```

And creating a script that automatically generates 25 songs into the playlist.
Changing the seed results in a completely new "sad" playlist. Check the playlist
out [here](https://open.spotify.com/playlist/3VDGx8xrXeoJlSgXfi5NQx?si=4fc9ec3beead4147).
It was definitely a success!
```{r, eval=FALSE}
# getting songs to be put in playlist as a vector of their uri:s
set.seed(2)
nordic_sad_tracks <- nordic_sad %>%
  sample_n(25) %>%
  pull(uri)

# getting id of the playlist in order to manipulate it
playlist_id <- get_my_playlists() %>%
  filter(name == "nordic_sad") %>%
  pull(id)

# removing all the current playlist tracks
current_playlist_tracks <- get_playlist_tracks(playlist_id = playlist_id)$track.uri
remove_tracks_from_playlist(playlist_id = playlist_id, uris = current_playlist_tracks)

# adding the new tracks to the playlist
add_tracks_to_playlist(playlist_id = playlist_id, uris = nordic_sad_tracks)
```

We can now move on to the next playlist, the international happy/upbeat one.
Since we are basically just repeating the same process, but with some few
tweaks, we leave the code for this out of the report.

```{r, include=FALSE}
international_happy_upbeat <- charts_features_df %>%
  select(uri, tempo, valence, energy) %>%
  distinct() %>%
  # centering aroung zero
  mutate(valence = valence - 0.5,
         energy = energy - 0.5) %>%
  # now only using tempos above 140 bpm
  filter(valence > 0.2 & energy > 0.2 & tempo > 140)
```

Here is a sample plot.

```{r, echo=FALSE, cache=TRUE}
set.seed(1)
international_happy_upbeat_plot <- international_happy_upbeat %>%
  sample_n(500) %>%
  ggplot(aes(x=valence, y=energy)) +
  geom_point(size = 0.3) +
  geom_vline(xintercept = 0) +
  geom_hline(yintercept = 0) +
  xlim(-0.5, 0.5) +
  ylim(-0.5, 0.5) +
  geom_label(x=-0.3, y=0.32, label = "Angry/Turbulent", color = "red") +
  geom_label(x=-0.3, y=-0.32, label = "Sad/Depressed", color = "blue") +
  geom_label(x=0.3, y=0.32, label = "Happy/Delighted", color = "green") +
  geom_label(x=0.3, y=-0.32, label = "Relaxed/Mellow", color = "purple") +
  ggtitle("Sample of (allegedly) happy songs popular in the world 2022 (n=500)")

international_happy_upbeat_plot
```

```{r, eval=FALSE, echo=FALSE}
# create the actual playlist using spotifyr
create_playlist(user_id = "31m5lue5uif6evna5thejrywzzgi",
                name = "international_happy_upbeat")
```

```{r, eval=FALSE, echo=FALSE}
# getting songs to be put in playlist as a vector of their uri:s
set.seed(5)
international_happy_upbeat_tracks <- international_happy_upbeat %>%
  sample_n(25) %>%
  pull(uri)

# getting id of the playlist in order to manipulate it
playlist_id <- get_my_playlists() %>%
  filter(name == "international_happy_upbeat") %>%
  pull(id)

# removing all the current playlist tracks
current_playlist_tracks <- get_playlist_tracks(playlist_id = playlist_id)$track.uri
remove_tracks_from_playlist(playlist_id = playlist_id, uris = current_playlist_tracks)

# adding the new tracks to the playlist
add_tracks_to_playlist(playlist_id = playlist_id, uris = international_happy_upbeat_tracks)
```

[Here](https://open.spotify.com/playlist/4RhNWRLFs85zlJp3KYwx2B?si=e8bf4f5c66c94225)
is the resulting playlist. It is definitely harder to call this a success. The
playlist feels too chaotic. This is probably due to the fact that we pulled songs
from all over the world. The tempos might be matching, but since there are so
many different languages, it is hard to find a common theme throughout the playlist.
In the case of a feeling such as happiness, it might be a better idea to have
some type of geographical restriction to the tracks' popularity.

# Conclusions and reflections

There are way more playlists that one can create using only the
valence, energy and tempo variables. The ones we have created in this report
only serve as an example as to how to work with the variables. For example,
one could try making the sampling more sophisticated than the sort of
"rectangular" sampling that we used. As an example, it would be interesting to
see what kind of results are yielded by exploring the borders between quadrants.

In addition to the variables we used for this project, there are many more -
check out the charts_features_df in order to get the full scope of the audio
features provided by Spotify.
If one were to incorporate some of these into the valence/energy mix, as well
as music genre and geopgraphical restrictions - it would probably be possible
to create playlists based on pretty specific requirements.

All in all, I hope this report served to inspire the readers' imagination of the
possibilities of automatically creating playlists in Spotify.

```{r, include=FALSE, eval=FALSE}
# saving plots for presentation purposes
ggsave("../Presentation/plots/example_plot.png", plot = example_plot,
       width = 8, height = 5)
ggsave("../Presentation/plots/nordic_sad_plot.png", plot = nordic_sad_plot,
       width = 8, height = 5)
ggsave("../Presentation/plots/international_happy_upbeat_plot.png",
       plot = international_happy_upbeat_plot,
       width = 8, height = 5)
```


# Sources
[spotifyr](https://www.rcharlie.com/spotifyr/)\
[exploring spotify with r tutorial](https://msmith7161.github.io/what-is-speechiness/)\
[RSelenium tutorial](http://joshuamccrain.com/tutorials/web_scraping_R_selenium.html)\
[Country codes and contient](https://gist.githubusercontent.com/stevewithington/20a69c0b6d2ff846ea5d35e5fc47f26c/raw/13716ceb2f22b5643ce5e7039643c86a0e0c6da6/country-and-continent-codes-list-csv.csv)\
[Spotify audio track features docs](https://developer.spotify.com/documentation/web-api/reference/#/operations/get-several-audio-features)\
