---
title: "The Right Spotify Playlist for the Occassion"
author: "August Jonasson (@aggelitoo)"
date: "2023-01-12"
output:
  ioslides_presentation:
    logo: SU_logo_CMYK.png
    incremental: TRUE
    css: slides.css
---

```{r setup, include=FALSE}
library(imager)
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction
- Scraping tracks from the Spotify top lists from all over the world
- Using Spotify API to enrich tracks with audio features
- Automatically generating playlists based on emotion to facilitate
- Creating the playlists in R through the Spotify API
- Presentation focus on audio features -> creating playlists

## Mood variables
- Valence (positive/negative)
- Energy (high/low)
- Tempo (beats per minute) - playlist coherence

## Combinations of valence/energy
- High valence, high energy: **Happy/delighted**
- High valence, low energy: **Relaxed/mellow**
- Low valence, low energy: **Sad/depressed**
- Low valence, high energy: **Angry/frustrated**

## Valence/energy

```{r, echo=FALSE}
example_songs_plot <- load.image("plots/example_plot.png")
plot(example_songs_plot)
```

## Trying to create a Sad Playlist

- Restricting the playlist by geograpical popularity (Nordic countries)
- Choosing songs with low valence/energy
- BPM below 100 (for coherence)

## Selection of songs
```{r cars, echo = FALSE}
nordic_sad_plot <- load.image("plots/nordic_sad_plot.png")
plot(nordic_sad_plot)
```

## Conclusions

- Geographical/cultural restriction contribute to coherence
- Holiday music (filtering)
- Many more variables and combinations to explore
- Thank you for listening!

